---
title: "Premier League Match Uncertainty as a Relegation Signal"
subtitle: "A Bayesian Hierarchical Analysis of Shannon Entropy from Bookmaker Odds"
author:
  - name: Naren Prakash
format:
  html:
    theme: cosmo
    css: "poster_style.css"
    toc: false
    number-sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 5)
library(tidyverse)
library(brms)
library(implied)
library(patchwork)
```

# Research Question

**Goal:** Determine whether match-level uncertainty (entropy) contains predictive signal for relegation outcomes, beyond what is captured by bookmaker probabilities alone.

**Hypothesis:** Higher average match uncertainty (entropy) for a team during the season is associated with a higher probability of relegation, controlling for other factors.

**Approach:** Construct entropy features from calibrated bookmaker odds for each team each season, then use a Bayesian hierarchical logistic regression to predict eventual relegation, pooling across teams and seasons.

# Data & Notation

- **Match** $i$ in **season** $s$ between home team $h$ and away team $a$
- **Power-transformed** Bet365 probabilities: $p_{i,s} = (p_{i,w}, p_{i,d}, p_{i,l})$
  - Buchdahl (2019) power / logarithmic method via `implied` R package
  - Alternative methods available: additive, Shin's, multiplicative
- **Relegation outcome**: $Y_{c,s} \in \{0, 1\}$ = 1 if team $c$ is relegated at end of season $s$

**Note:** The model uses only information available *during* the season to predict an outcome realized *after* the season ends.

| Detail | Value |
|--------|-------|
| Source | Bet365 odds, English Premier League |
| Seasons | 2004–2025 (22 seasons) |

# Shannon Entropy

For match $i$ in season $s$:

$$H_{i,s} = -\sum_{k \in \{w,d,l\}} p_{i,s,k} \log_3(p_{i,s,k})$$

Base-3 log normalizes entropy to $[0, 1]$: 0 = perfectly predictable, 1 = maximum uncertainty.

# Team-Level Entropy Features

For each team $c$ in each season $s$:

| Feature | Definition |
|---------|-----------|
| Mean entropy | $\bar{H}_{c,s} = \frac{1}{N}\sum_i H_{i,s}$ |
| Entropy variance | $\text{Var}(H)_{c,s} = \frac{1}{N-1}\sum_i (H_{i,s} - \bar{H}_{c,s})^2$ |

# Bayesian Hierarchical Model

$$Y_{c,s} \sim \text{Bernoulli}(p_{c,s})$$

$$\text{logit}(p_{c,s}) = \alpha + \beta_1 \cdot \bar{H}_{c,s} + \beta_2 \cdot \text{Var}(H)_{c,s} + \gamma_s$$

$$\gamma_s \sim \mathcal{N}(0, \sigma_{\text{season}})$$

**Priors** (predictors standardized to z-scores):

| Parameter | Prior | Justification |
|-----------|-------|---------------|
| $\beta_1, \beta_2$ | $\mathcal{N}(0, 1.5)$ | Weakly informative; 1.5 on logit shifts probability from ~15% to ~50% |
| $\sigma_{\text{season}}$ | $\mathcal{N}^+(0, 1)$ | Half-normal; modest season-to-season variation expected |
| $\alpha$ | $\mathcal{N}(0, 2)$ | Covers 2%–65% relegation probability at 95% (true rate = 15%) |

# Data Summary

```{r}
games <- read_csv("games.csv", show_col_types = FALSE)
season_files <- list.files("Seasonal Data", pattern = "Season02.*\\.csv$", full.names = TRUE)

odds_input <- games %>% select(B365H, B365D, B365A)
transformed <- implied_probabilities(odds_input, method = "power")
prob_df <- data.frame(transformed$probabilities) %>%
  rename(p_home = B365H, p_draw = B365D, p_away = B365A)

games_proc <- games %>%
  bind_cols(prob_df) %>%
  mutate(
    entropy = -(p_home * log(p_home, base = 3) +
                  p_draw * log(p_draw, base = 3) +
                  p_away * log(p_away, base = 3))
  )
```

```{r, fig.height=4}
ggplot(games_proc, aes(x = entropy)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = "#2C3E50", color = "white", alpha = 0.8) +
  geom_density(color = "#E74C3C", linewidth = 1.2) +
  geom_vline(aes(xintercept = mean(entropy)),
             linetype = "dashed", color = "#E67E22", linewidth = 1.2) +
  labs(
    title = "Match-Level Entropy Distribution",
    subtitle = sprintf("n = %d matches | mean = %.3f",
                       nrow(games_proc), mean(games_proc$entropy)),
    x = "Shannon Entropy (base 3)", y = "Density"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 14),
        axis.title = element_text(size = 12), axis.text = element_text(size = 11))
```

# Mean-Variance Profile

```{r}
compute_standings <- function(season_num) {
  sf <- season_files[str_detect(season_files, sprintf("%04d", season_num))]
  if (length(sf) == 0 || !file.exists(sf)) return(NULL)
  s <- read_csv(sf, show_col_types = FALSE)
  h <- s %>% select(team = HomeTeam, FTHG, FTAG, FTR) %>%
    mutate(pts = case_when(FTR == "H" ~ 3, FTR == "D" ~ 1, TRUE ~ 0), gd = FTHG - FTAG)
  a <- s %>% select(team = AwayTeam, FTHG, FTAG, FTR) %>%
    mutate(pts = case_when(FTR == "A" ~ 3, FTR == "D" ~ 1, TRUE ~ 0), gd = FTAG - FTHG)
  bind_rows(h, a) %>% group_by(team) %>%
    summarise(final_pts = sum(pts), final_gd = sum(gd), .groups = "drop") %>%
    arrange(desc(final_pts), desc(final_gd)) %>%
    mutate(pos = row_number(), relegated = pos >= 18, season = season_num)
}

standings <- map_dfr(2004:2025, compute_standings)

ent_feat <- games_proc %>%
  select(team = HomeTeam, Season, entropy) %>%
  bind_rows(games_proc %>% select(team = AwayTeam, Season, entropy)) %>%
  group_by(team, Season) %>%
  summarise(mean_H = mean(entropy), var_H = var(entropy), .groups = "drop")

md <- standings %>%
  left_join(ent_feat, by = c("team" = "team", "season" = "Season")) %>%
  drop_na(mean_H, var_H)

sample_seasons <- c(2004, 2013, 2023)
md_sample <- md %>% filter(season %in% sample_seasons) %>%
  mutate(
    zone = case_when(
      pos <= 4 ~ "Champions League", pos <= 6 ~ "Europa League",
      pos <= 17 ~ "Mid-table", TRUE ~ "Relegation Zone"
    ),
    zone = factor(zone, levels = c("Champions League", "Europa League", "Mid-table", "Relegation Zone"))
  )
```

```{r, fig.height=5}
ggplot(md_sample, aes(x = mean_H, y = var_H)) +
  geom_point(aes(color = zone), alpha = 0.8, size = 3.5, stroke = 2) +
  geom_text(aes(label = team, color = zone), size = 3.5, vjust = -1, hjust = 0.5, fontface = "bold") +
  scale_color_manual(name = "Zone", values = c(
    "Champions League" = "#27AE60", "Europa League" = "#2ECC71",
    "Mid-table" = "#3498DB", "Relegation Zone" = "#E74C3C"
  )) +
  facet_wrap(~season, ncol = 3) +
  labs(title = "Entropy Mean-Variance Profile (Sample Seasons)",
       x = "Mean Entropy", y = "Entropy Variance") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 14),
        axis.title = element_text(size = 12), axis.text = element_text(size = 11),
        legend.position = "bottom", strip.text = element_text(face = "bold", size = 12))
```

# Relegated vs Safe Teams

```{r, fig.height=5}
md_long <- md_sample %>%
  mutate(relegated_label = ifelse(relegated, "Relegated", "Safe"),
         relegated_label = factor(relegated_label, levels = c("Safe", "Relegated"))) %>%
  pivot_longer(
    cols = c(mean_H, var_H),
    names_to = "metric", values_to = "value"
  ) %>%
  mutate(metric = case_when(
    metric == "mean_H" ~ "Mean Entropy", metric == "var_H" ~ "Entropy Variance"
  ))

ggplot(md_long, aes(x = relegated_label, y = value, fill = relegated_label)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA, linewidth = 1) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 2) +
  facet_grid(metric ~ season, scales = "free_y") +
  scale_fill_manual(name = "", values = c("Safe" = "#3498DB", "Relegated" = "#E74C3C")) +
  labs(title = "Entropy Features: Relegated vs Safe Teams", x = "", y = "") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 14),
        axis.text = element_text(size = 11), legend.position = "none",
        strip.text = element_text(face = "bold", size = 11))
```

# Posterior Results

```{r}
fit <- readRDS("fit_bayesian_entropy_relegation.rds")
post <- as_draws_df(fit)
```

```{r, out.width="70%"}
coef_df <- tibble(
  Parameter = c("Mean Entropy", "Entropy Variance", "Intercept"),
  Median = c(median(post$b_mean_entropy_z), median(post$b_var_entropy_z), median(post$b_Intercept)),
  `2.5%` = c(quantile(post$b_mean_entropy_z, 0.025), quantile(post$b_var_entropy_z, 0.025), quantile(post$b_Intercept, 0.025)),
  `97.5%` = c(quantile(post$b_mean_entropy_z, 0.975), quantile(post$b_var_entropy_z, 0.975), quantile(post$b_Intercept, 0.975)),
  `P(>0)` = c(mean(post$b_mean_entropy_z > 0), mean(post$b_var_entropy_z > 0), mean(post$b_Intercept > 0))
)

knitr::kable(coef_df, digits = 3,
             col.names = c("Parameter", "Median", "2.5%", "97.5%", "P(>0)"),
             caption = "Posterior Summaries (standardized predictors)",
             booktabs = TRUE)
```

# Posterior Distributions

```{r, fig.height=5.5}
all_long <- post %>%
  select(b_mean_entropy_z, b_var_entropy_z, b_Intercept) %>%
  pivot_longer(everything(), names_to = "parameter", values_to = "value") %>%
  mutate(
    parameter = case_when(
      parameter == "b_mean_entropy_z" ~ "Mean Entropy",
      parameter == "b_var_entropy_z" ~ "Entropy Variance",
      parameter == "b_Intercept" ~ "Intercept"
    ),
    parameter = factor(parameter, levels = c("Intercept", "Mean Entropy", "Entropy Variance"))
  )

cs <- all_long %>% group_by(parameter) %>%
  summarise(med = median(value), lo = quantile(value, 0.025), hi = quantile(value, 0.975), .groups = "drop")

p1 <- ggplot(all_long %>% filter(parameter == "Mean Entropy"), aes(x = value)) +
  geom_histogram(bins = 50, fill = "#3498DB", color = "white", alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", linewidth = 1) +
  geom_vline(xintercept = cs$med[cs$parameter == "Mean Entropy"], color = "black", linewidth = 1.5) +
  labs(title = "Mean Entropy", x = "Coefficient (log-odds)", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 14),
        axis.title = element_text(size = 11), axis.text = element_text(size = 10))

p2 <- ggplot(all_long %>% filter(parameter == "Entropy Variance"), aes(x = value)) +
  geom_histogram(bins = 50, fill = "#E67E22", color = "white", alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", linewidth = 1) +
  geom_vline(xintercept = cs$med[cs$parameter == "Entropy Variance"], color = "black", linewidth = 1.5) +
  labs(title = "Entropy Variance", x = "Coefficient (log-odds)", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 14),
        axis.title = element_text(size = 11), axis.text = element_text(size = 10))

p3 <- ggplot(all_long %>% filter(parameter == "Intercept"), aes(x = value)) +
  geom_histogram(bins = 50, fill = "#2C3E50", color = "white", alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", linewidth = 1) +
  geom_vline(xintercept = cs$med[cs$parameter == "Intercept"], color = "black", linewidth = 1.5) +
  labs(title = "Intercept", x = "Coefficient (log-odds)", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 14),
        axis.title = element_text(size = 11), axis.text = element_text(size = 10))

p1 / p2 / p3 + plot_layout(heights = c(1, 1, 1))
```

# Prior Sensitivity Analysis

```{r}
fit_main <- readRDS("fit_bayesian_entropy_relegation.rds")
fit_flat <- readRDS("fit_sensitivity_default.rds")
fit_skep <- readRDS("fit_sensitivity_skeptical.rds")

extract_coefs <- function(fit) {
  post2 <- as_draws_df(fit)
  tibble(
    Median_mean = median(post2$b_mean_entropy_z), Median_var = median(post2$b_var_entropy_z),
    Lower_mean = quantile(post2$b_mean_entropy_z, 0.025), Lower_var = quantile(post2$b_var_entropy_z, 0.025),
    Upper_mean = quantile(post2$b_mean_entropy_z, 0.975), Upper_var = quantile(post2$b_var_entropy_z, 0.975)
  )
}

sens_df <- bind_rows(
  extract_coefs(fit_main) %>% mutate(Prior = "Weakly Informative"),
  extract_coefs(fit_flat) %>% mutate(Prior = "Flat (Default)"),
  extract_coefs(fit_skep) %>% mutate(Prior = "Skeptical")
) %>% mutate(Prior = factor(Prior, levels = c("Weakly Informative", "Flat (Default)", "Skeptical"))) %>%
  pivot_longer(
    cols = c(Median_mean, Median_var, Lower_mean, Lower_var, Upper_mean, Upper_var),
    names_pattern = "(.*)_(mean|var)", names_to = c(".value", "param"), values_to = "value"
  ) %>% mutate(Parameter = ifelse(param == "mean", "Mean Entropy", "Entropy Variance"))
```

```{r, fig.height=5}
ggplot(sens_df, aes(x = Median, y = Prior)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 1) +
  geom_errorbarh(aes(xmin = Lower, xmax = Upper, color = Prior), height = 0.3, linewidth = 1.5) +
  geom_point(aes(color = Prior), size = 4.5) +
  facet_wrap(~Parameter, scales = "free_x", ncol = 1) +
  scale_color_viridis_d(option = "plasma", end = 0.8) +
  labs(title = "Robustness Across Prior Choices",
       subtitle = "Posterior medians and 95% credible intervals",
       x = "Coefficient (log-odds)", y = "") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 14),
        axis.text = element_text(size = 12), axis.title = element_text(size = 12),
        legend.position = "none", strip.text = element_text(face = "bold", size = 12))
```

**Weakly Informative**: $\mathcal{N}(0, 1.5)$ — allows moderate effects. **Flat**: $\mathcal{N}(0, 10)$ — nearly non-informative; tests if data alone drives inference. **Skeptical**: $\mathcal{N}(0, 0.5)$ — encodes null that entropy has negligible predictive power; ~10pp change per SD.

# Season-Level Variation

```{r, fig.height=4}
se <- ranef(fit_main)$season[, , "Intercept"] %>%
  as.data.frame() %>%
  setNames(c("Estimate", "Est.Error", "Q2.5", "Q97.5")) %>%
  mutate(season = as.integer(rownames(.))) %>%
  arrange(season)

ggplot(se, aes(x = season, y = Estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", linewidth = 1) +
  geom_col(fill = "#2C3E50", alpha = 0.8) +
  geom_errorbar(aes(ymin = Q2.5, ymax = Q97.5), width = 0.5, color = "#E74C3C", linewidth = 1) +
  scale_x_continuous(breaks = seq(2004, 2025, 2)) +
  labs(title = "Season-Level Varying Intercepts",
       subtitle = "Estimates with 95% credible intervals",
       x = "Season", y = "Intercept (log-odds)") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 14),
        axis.title = element_text(size = 12), axis.text = element_text(size = 11))
```

# Future Work

1. **In-season prediction:** Compute rolling entropy up to a fixed matchweek (e.g., MW 30) for mid-season relegation forecasting rather than post-hoc analysis.

2. **Structured priors:** Explore Cauchy(0, 2.5) (Gelman et al., 2008) and R2-D2 shrinkage (Zhang et al., 2022) for principled regularization.

# References

Buchdahl (2019), *Wisdom of the Crowds*.
Clarke et al. (2017), *Am. J. Sports Sci., 5*(6).
